Tensor denoising and completion based on ordinal observations

### 2020-12-02 paper/arxiv_ordinal/arxiv.tex
Add files via upload

%%%%%%%% ICML 2020 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables
\usepackage{caption}
% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2020} with \usepackage[nohyperref]{icml2020} above.
\usepackage{hyperref}

@@

% If accepted, instead use the following line for the camera-ready submission:
\usepackage[accepted]{icml2020}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Ordinal tensor denoising and completion}
\usepackage{wrapfig}
\usepackage{multirow}
\usepackage{graphicx}
%\usepackage[utf8]{inputenc} % allow utf-8 input
%\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
%\usepackage{booktabs}       % professional-quality tables
\usepackage{amsmath,amssymb}
\usepackage{amsthm}    % blackboard math symbols
%\usepackage{nicefrac}       % compact symbols for 1/2, etc.
%\usepackage{microtype}      % microtypography
\usepackage{bm}
%\usepackage{subfig}
%\usepackage[english]{babel}
%\usepackage{algorithm}
%\usepackage{appendix}
\usepackage{mathtools}
\mathtoolsset{showonlyrefs}
\usepackage{enumitem}
\theoremstyle{plain}
\newtheorem{thm}{Theorem}[section]
\newtheorem{lem}{Lemma}
\newtheorem{prop}{Proposition}
\newtheorem{pro}{Property}
\newtheorem{assumption}{Assumption}

@@

@@

\usepackage{amssymb}
\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%

\begin{document}

\twocolumn[
\icmltitle{Tensor denoising and completion based on ordinal observations}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2020
% package.

% List of affiliations: The first argument should be a (short)
% identifier you will use later to specify author affiliations
% Academic affiliations should list Department, University, City, Region, Country
% Industry affiliations should list Company, City, Region, Country

% You can specify symbols, otherwise they are numbered in order.
% Ideally, you should not use this facility. Affiliations will be numbered
% in order of appearance and this is the preferred way.
%\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Chanwoo Lee}{wi}
\icmlauthor{Miaoyan Wang}{wi}
\end{icmlauthorlist}

\icmlaffiliation{wi}{Department of Statistics, University of Wisconsin-Madison, Wisconsin, USA}
\icmlcorrespondingauthor{Miaoyan Wang}{miaoyan.wang@wisc.edu}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Higher-order tensors, ordinal observation, tensor decomposition, tensor completion}

\vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column
% listing the affiliations and the copyright notice.
% The command takes one argument, which is text to display at the start of the footnote.
% The \icmlEqualContribution command is standard text for equal contribution.
% Remove it (just {}) if you do not need this facility.

\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
%\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.

\begin{abstract}
Higher-order tensors arise frequently in applications such as neuroimaging, recommendation system, and social network analysis. We consider the problem of low-rank tensor estimation from possibly incomplete, ordinal-valued observations. Two related problems are studied, one on tensor denoising and another on tensor completion. We propose a multi-linear cumulative link model, develop a rank-constrained M-estimator, and obtain theoretical accuracy guarantees. Our mean squared error bound enjoys a faster convergence rate than previous results, and we show that the proposed estimator is minimax optimal under the class of low-rank models. Furthermore, the procedure developed serves as an efficient completion method which guarantees consistent recovery of an order-$K$ $(d,\ldots,d)$-dimensional low-rank tensor using only $\tilde \tO(Kd)$ noisy, quantized observations. We demonstrate the outperformance of our approach over previous methods on the tasks of clustering and collaborative filtering.

@@

\begin{table*}[ht]
\centerline{\includegraphics[width=1\textwidth]{compare.pdf}}
\caption{Comparison with previous work. For ease of presentation, we summarize the error rate and sample complexity assuming equal tensor dimension in all modes. $K$: tensor order; $L$: number of ordinal levels; $d$: dimension at each mode. }\label{tab:compare}
\end{table*}

@@

\begin{example}[Logistic model] The logistic model is characterized by~\eqref{eq:model} with $f(\theta)=(1+e^{-\theta/\sigma})^{-1}$, where $\sigma>0$ is the scale parameter. Equivalently, the noise $\varepsilon_\omega$ in~\eqref{eq:quantization} follows i.i.d.\ logistic distribution with scale parameter $\sigma$.
\end{example}
\begin{example}[Probit model] The probit model is characterized by~\eqref{eq:model} with
$f(\theta)=\mathbb{P}(z\leq \theta/\sigma)$, where $z\sim N(0,1)$. Equivalently, the noise $\varepsilon_\omega$ in~\eqref{eq:quantization} follows i.i.d.\ $N(0,\sigma^2)$.
\end{example}
Other link functions are also possible, such as Laplace, Cauchy, etc~\cite{mccullagh1980regression}. All the models share the property that the ordered categories can be thought of as contiguous interval on some continuous scale. We should point out that, although the latent-variable interpretation is incisive, our estimation procedure does not refer to the existence of $\tY^*$. Therefore, our model~\eqref{eq:model} is general and still valid in the absence of quantization process. More generally, we make the following assumptions about the link $f$.

\begin{assumption}\label{ass:link}
The link function is assumed to satisfy:
\vspace{-.2cm}
\begin{enumerate}
\item $f(\theta)$ is strictly increasing and twice-differentiable in $\theta\in \mathbb{R}$.
\item $f'(\theta)$ is strictly log-concave and symmetric with respect to $\theta=0$.
\end{enumerate}
\end{assumption}

@@

Clearly, the problem (P1) cannot be solved uniformly for all possible $\Theta$. We focus on a class of ``low-rank'' and ``flat'' signal tensors, which is a plausible assumption in practical applications~\cite{zhou2013tensor,bhaskar20151}. Specifically, we consider the parameter space:
\begin{equation}\label{eq:space}
\tP=\left\{\Theta\in\mathbb{R}^{d_1\times \cdots \times d_K} \colon \text{rank}(\Theta)\leq \mr,\ \mnormSize{}{\Theta}\leq \alpha\right\}.
\end{equation}
where $\mr=(r_1,\ldots,r_K)$ denotes the Tucker rank of $\Theta$.

The parameter tensor of our interest satisfies two constraints. The first is that $\Theta$ is a low-rank tensor, with $r_k=\tO(1)$ for all $k\in[K]$. Equivalently, $\Theta$ admits the Tucker decomposition:
\begin{equation}\label{eq:Tucker}
\Theta=\tC\times_1\mM_1\times_1 \cdots \times_K \mM_K,
\end{equation}
where $\tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ is a core tensor, $\mM_k\in\mathbb{R}^{d_k\times r_k}$ are factor matrices with orthogonal columns, and $\times_k$ denotes the tensor-by-matrix multiplication~\cite{kolda2009tensor}. The Tucker low-rankness is popularly imposed in tensor data analysis, and is shown to provide a reasonable tradeoff between model complexity and model flexibility. Note that, unlike matrices, there are various notations of tensor low-rankness, such as CP rank~\cite{hitchcock1927expression} and train rank~\cite{oseledets2011tensor}. Some notation of low-rankness may lead to mathematically ill-posed optimization; for example, the best low CP-rank tensor approximation may not exist~\cite{de2008tensor}. We choose Tucker representation for well-posedness of optimization and easy interpretation.

The second constraint is that the entries of $\Theta$ are uniformly bounded in magnitude by a constant $\alpha \in \mathbb{R}_{+}$. In view of~\eqref{eq:latent}, we refer to $\alpha$ as the signal level. The entry-wise bound assumption is a technical condition that avoids the degeneracy in probability estimation with ordinal observations.

@@

\section{Numerical Implementation}\label{sec:algorithm}
We describe the algorithm to seek the optimizer of~\eqref{eq:objective}. In practice, the cut-off points $\mb$ are often unknown, so we choose to maximize $\logl$ jointly over $(\Theta,\mb)\in\tP\times \tB$ (see the Supplement for details). The objective $\logl$ is concave in $(\Theta,\mb)$ whenever $f'$ is log-concave. However, the feasible set $\tP$ is non-convex, which makes the optimization~\eqref{eq:objective} a non-convex problem. We employ the alternating optimization approach by utilizing the Tucker representation of $\Theta$. Specifically, based on~\eqref{eq:Tucker} and~\eqref{eq:objective}, the objective function consists of $K+2$ blocks of variables, one for the cut-off points $\mb$, one for the core tensor $\tC$, and $K$ for the factor matrices $\mM_k$'s. The optimization is a simple convex problem if any $K+1$ out of the $K+2$ blocks are fixed. We update one block at a time while holding others fixed, and alternate the optimization throughout the iteration. The convergence is guaranteed whenever $\logl$ is bounded from above, since the alternating procedure monotonically increases the objective. The Algorithm~\ref{alg} gives the full description.

@@

\begin{figure}[H]
\centering
\includegraphics[width=8cm]{algorithm.pdf}
\caption{Trajectory of objective function with various $d$ and $r$.}\label{fig:stability}
\end{figure}

Second, the algorithm takes the rank $\mr$ as an input. In practice, the rank $\mr$ is hardly known and needs to be estimated from the data. We suggest to use Bayesian information criterion (BIC) and choose the rank that minimizes BIC; i.e.
\begin{align}\label{eq:BIC}
\hat \mr&=\argmin_{\mr\in\mathbb{N}^K_{+}} \text{BIC}(\mr)\\
&=\argmin_{\mr\in\mathbb{N}^K_{+}}\{-2\tL_{\tY}(\hat \Theta(\mr),\hat \mb(\mr))+p_e(\mr)\log (\prod_k d_k)\},
\end{align}
where $\hat \Theta(\mr), \hat\mb(\mr)$ are the estimates given the rank $\mr$, and $p_e(\mr)\stackrel{\text{def}}{=}\sum_k (d_k-r_k)r_k+\prod_k r_k$ is the effective number of parameters in the model. We select $\hat \mr$ that minimizes BIC through a grid search. The choice of BIC is intended to balance between the goodness-of-fit for the data and the degrees of freedom in the population model.

\section{Experiments}\label{sec:experiment}
In this section, we evaluate the empirical performance of our methods\footnote{Software package: \url{https://CRAN.R-project.org/package=tensorordinal}}. We investigate both the complete and the incomplete settings, and compare the recovery accuracy with other tensor-based methods. Unless otherwise stated, the ordinal data tensors are generated from model~\eqref{eq:model} using standard probit link $f$. We consider the setting with $K=3$, $d_1=d_2=d_3=d$, and $r_1=r_2=r_3=r$. The parameter tensors are simulated based on~\eqref{eq:Tucker}, where the core tensor entries are i.i.d.\ drawn from $\tN(0,1)$, and the factors $\mM_k$ are uniformly sampled (with respect to Haar measure) from matrices with orthonormal columns. We set the cut-off points $b_\ell=f^{-1}(\ell/L)$ for $\ell\in[L]$, such that $f(b_\ell)$ are evenly spaced from 0 to 1. In each simulation study, we report the summary statistics across $n_{\text{sim}}$ = 30 replications.

@@

\begin{figure}[http]
\includegraphics[width=8.3cm]{panel.pdf}
\caption{Empirical relationship between (relative) MSE versus (a) dimension $d$, (b) signal level $\alpha$, (c) observation fraction $\rho$, and (d) number of ordinal levels $L$. In panels (b)-(d), we plot the relative MSE $=\FnormSize{}{\hat \Theta-\trueT}^2/\FnormSize{}{\trueT}^2$ for better visualization.}
\label{fig:finite}
\vspace{-.3cm}
\end{figure}

\vspace{-.2cm}
\subsection{Comparison with alternative methods}
Next, we compare our ordinal tensor method ({\bf Ordinal-T}) with three popular low-rank methods:
\begin{itemize}[itemsep=0pt,topsep=0pt,leftmargin=*,partopsep=0pt]
\item Continuous tensor decomposition ({\bf Continuous-T})~\cite{acar2010scalable} is a low-rank approximation method based on classical Tucker model.
\item One-bit tensor completion ({\bf 1bit-T})~\cite{ghadermarzy2018learning} is a max-norm penalized tensor learning method based on partial binary observations.
\item Ordinal matrix completion ({\bf Ordinal-M})~\cite{bhaskar2016probabilistic} is a rank-constrained matrix estimation method based on noisy, quantized observations.
\end{itemize}

We apply each of the above methods to $L$-level ordinal tensors $\tY$ generated from model~\eqref{eq:model}. The {\bf Continuous-T} is applied directly to $\tY$ by treating the $L$ levels as continuous observations. The {\bf Ordinal-M} is applied to the 1-mode unfolding matrix $\tY_{(1)}$. The {\bf 1bit-T} is applied to $\tY$ in two ways. The first approach ({\bf 1bit-sign-T}) follows from~\citet{ghadermarzy2018learning} that transforms
$\tY$ to a binary tensor, by taking the entrywise sign of the mean-adjusted tensor, $\tY- |\Omega|^{-1}\sum_\omega y_\omega$. The second approach ({\bf 1bit-category-T}) transforms the order-3 ordinal tensor $\tY$ to an order-4 binary tensor, $\tY^{\sharp}=\entry{y^{\sharp}_{ijkl}}$, via dummy variable encoding; i.e., $y^{\sharp}_{ijk\ell}=\mathds{1}\{y_{ijk}=\ell\}$ for $\ell\in[L-1]$. We evaluate the methods by their capabilities in predicting the most likely labels, $y_\omega^{\text{mode}}=\arg\max_\ell\mathbb{P}(y_\omega=\ell)$. Two performance metrics are considered: mean absolute deviation (MAD) $=d^{-K}\sum_\omega |y_\omega^{\text{mode}}-\hat y_\omega^{\text{mode}}|$, and misclassification rate (MCR) $=d^{-K}\sum_\omega\mathds{1}\{y_\omega^{\text{mode}}\neq\text{round}(\hat y_\omega^{\text{mode}})\}$,
where $\text{round}(\cdot)$ denotes the nearest integer. Note that MAD penalizes the large deviation more heavily than MCR.

@@

\begin{figure}[http]
\includegraphics[width=8.3cm]{panel2.pdf}
\caption{Performance comparison for predicting most likely labels. (a, c) Prediction errors versus sample complexity $\rho=|\Omega|/d^K$ when $L=5$. (b, d) Prediction errors versus the number of ordinal levels $L$ when $\rho=0.8$. }\label{fig:compare}
\vspace{-.4cm}
\end{figure}

\begin{figure}[http]
\includegraphics[width=8.3cm]{panel3.pdf}
\caption{Performance comparison for predicting median labels. (a, c) Prediction error versus sample complexity $\rho=|\Omega|/d^K$ when $L=5$. (b, d) Prediction error versus the number of ordinal levels $L$, when $\rho=0.8.$}\label{fig:compare2}
\vspace{-.4cm}
\end{figure}

We also compare the methods by their performance in predicting the median labels, $y_\omega^{\text{median}}=\min\{\ell\colon\mathbb{P}(y_\omega=\ell)\geq 0.5\}$. Under the latent variable model~\eqref{eq:latent} and Assumption~\ref{ass:link}, the median label is the quantized $\theta_\omega$ without noise; i.e.\ $y_\omega^{\text{median}}=\sum_\ell \mathds{1}\{\theta_\omega\in(b_{\ell-1},b_\ell]\}$.
 We utilize the same simulation setting as in the earlier experiment. Figure~\ref{fig:compare2} shows that our method outperforms the others in both MCR and MAD. The improved accuracy comes from the incorporation of multilinear low-rank structure, multi-level observations, and the ordinal structure. Interestingly, the median estimator tends to yield smaller MAD than the mode estimator, $\text{MAD}(\tY^{\text{median}}, \hat \tY^{\text{median}})\leq \text{MAD}(\tY^{\text{mode}}, \hat \tY^{\text{mode}})$  (Figures~\ref{fig:compare}a-b vs.\ Figures~\ref{fig:compare2}a-b), for the three multilevel methods ({\bf 1bit-sign-T}, {\bf Ordinal-M}, and {\bf Ordinal-T}). The mode estimator, on the other hand, tends to yield smaller MCR than the median estimator, $\text{MCR}(\tY^{\text{mode}}, \hat \tY^{\text{mode}})\leq\text{MCR}(\tY^{\text{median}}, \hat \tY^{\text{median}})$ (Figures~\ref{fig:compare}c-d vs.\ Figures~\ref{fig:compare2}c-d). This tendency is from the property that the median estimator $\hat y^{(\text{median})}_\omega$ minimizes $R_1(z)=\mathbb{E}_{y_\omega}|y_\omega-z|$, whereas the mode estimator $\hat y^{(\text{mode})}_\omega $ minimizes $R_2(z)=\mathbb{E}_{y_\omega}\mathds{1}\{y_\omega=z\}$. Here the expectation is over the categorical distribution of $y_\omega$ given parameters $\hat \Theta$ and $\hat \mb$.

\section{Data Applications}
We apply our ordinal tensor method to two real-world datasets. In the first application, we use our model to analyze an ordinal tensor consisting of structural connectivities among 68 brain regions for 136 individuals from Human Connectome Project (HCP)~\citep{geddes2016human}. In the second application, we perform tensor completion to an ordinal dataset with missing values. The data tensor records the ratings of 139 songs on a scale of 1 to 5 from 42 users on 26 contexts~\citep{baltrunas2011incarmusic}.

\begin{table*}[http]
\includegraphics[width=1\textwidth]{HCPtable.pdf}
\vspace{-.7cm}
\caption{Node clusters in the HCP analysis. The first alphabet in the node name indicates the left (L) or right (R) hemisphere. The number in the parentheses indicates the node count in each cluster. }  \label{table:clustering}
\vspace{-.4cm}
\end{table*}

@@

\bibliography{tensor_wang}
\bibliographystyle{icml2020}

\clearpage
\appendix
\onecolumn

\icmltitle{Supplements for ``Tensor denoising and completion based on \\ ordinal observations''}

>>>>

\documentclass[11pt]{article}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{caption}
% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2020} with \usepackage[nohyperref]{icml2020} above.
\usepackage{hyperref}
\usepackage[margin=1in]{geometry}

@@

% If accepted, instead use the following line for the camera-ready submission:

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\usepackage{graphicx}
%\usepackage[utf8]{inputenc} % allow utf-8 input
%\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
%\usepackage{booktabs}       % professional-quality tables
\usepackage{amsmath,amssymb}
\usepackage{amsthm}    % blackboard math symbols
%\usepackage{nicefrac}       % compact symbols for 1/2, etc.
%\usepackage{microtype}      % microtypography
\usepackage{bm}
%\usepackage{subfig}
%\usepackage[english]{babel}
%\usepackage{algorithm}
%\usepackage{appendix}
\usepackage{mathtools}
\mathtoolsset{showonlyrefs}
\usepackage{enumitem}
\theoremstyle{plain}
\newtheorem{thm}{Theorem}[section]
\newtheorem{lem}{Lemma}
\newtheorem{prop}{Proposition}
\newtheorem{pro}{Property}
\newtheorem{assumption}{Assumption}

@@

% Use fancyhdr package
\usepackage{color}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{natbib}
\usepackage{forloop}

@@

\usepackage{amssymb}

\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%

\title{Tensor denoising and completion based on ordinal observations}
\date{}
\author{%
Chanwoo Lee \\
University of Wisconsin -- Madison\\
\texttt{chanwoo.lee@wisc.edu} \\
\and
Miaoyan Wang \\
University of Wisconsin -- Madison\\
\texttt{miaoyan.wang@wisc.edu} \\
}

\begin{document}
\maketitle
\begin{abstract}
Higher-order tensors arise frequently in applications such as neuroimaging, recommendation system, and social network analysis. We consider the problem of low-rank tensor estimation from possibly incomplete, ordinal-valued observations. Two related problems are studied, one on tensor denoising and another on tensor completion. We propose a multi-linear cumulative link model, develop a rank-constrained M-estimator, and obtain theoretical accuracy guarantees. Our mean squared error bound enjoys a faster convergence rate than previous results, and we show that the proposed estimator is minimax optimal under the class of low-rank models. Furthermore, the procedure developed serves as an efficient completion method which guarantees consistent recovery of an order-$K$ $(d,\ldots,d)$-dimensional low-rank tensor using only $\tilde \tO(Kd)$ noisy, quantized observations. We demonstrate the outperformance of our approach over previous methods on the tasks of clustering and collaborative filtering.

@@

\begin{table*}[ht]
\centerline{\includegraphics[width=1\textwidth]{compare.pdf}}
\caption{Comparison with previous work. For ease of presentation, we summarize the error rate and sample complexity assuming equal tensor dimension in all modes. $K$: tensor order; $L$: number of ordinal levels; $d$: dimension at each mode. }\label{tab:compare}

\end{table*}

@@

\begin{example}[Logistic model] The logistic model is characterized by~\eqref{eq:model} with $f(\theta)=(1+e^{-\theta/\sigma})^{-1}$, where $\sigma>0$ is the scale parameter. Equivalently, the noise $\varepsilon_\omega$ in~\eqref{eq:quantization} follows i.i.d.\ logistic distribution with scale parameter $\sigma$.
\end{example}
\begin{example}[Probit model] The probit model is characterized by~\eqref{eq:model} with
$f(\theta)=\mathbb{P}(z\leq \theta/\sigma)$, where $z\sim N(0,1)$. Equivalently, the noise $\varepsilon_\omega$ in~\eqref{eq:quantization} follows i.i.d.\ $N(0,\sigma^2)$.
\end{example}
Other link functions are also possible, such as Laplace, Cauchy, etc~\cite{mccullagh1980regression}. These latent variable models share the property that the ordered categories can be thought of as contiguous intervals on some continuous scale. We should point out that, although the latent-variable interpretation is incisive, our estimation procedure does not refer to the existence of $\tY^*$. Therefore, our model~\eqref{eq:model} is general and still valid in the absence of quantization process. More generally, we make the following assumptions about the link $f$.

\begin{assumption}\label{ass:link}
The link function is assumed to satisfy:

\begin{enumerate}
\item $f(\theta)$ is strictly increasing and twice-differentiable in $\theta\in \mathbb{R}$.
\item $f'(\theta)$ is strictly log-concave and symmetric with respect to $\theta=0$.
\end{enumerate}
\end{assumption}

@@

Clearly, the problem (P1) cannot be solved uniformly for all possible $\Theta$ with no assumptions. We focus on a class of ``low-rank'' and ``flat'' signal tensors, which is a plausible assumption in practical applications~\cite{zhou2013tensor,bhaskar20151}. Specifically, we consider the parameter space,
\begin{equation}\label{eq:space}
\tP=\left\{\Theta\in\mathbb{R}^{d_1\times \cdots \times d_K} \colon \text{rank}(\Theta)\leq \mr,\ \mnormSize{}{\Theta}\leq \alpha\right\},
\end{equation}
where $\mr=(r_1,\ldots,r_K)$ denotes the Tucker rank of $\Theta$.

The parameter tensor of our interest satisfies two constraints. The first is that $\Theta$ is a low-rank tensor, with $r_k=\tO(1)$ as $d_{\min}\to \infty$ for all $k\in[K]$. Equivalently, $\Theta$ admits the Tucker decomposition:
\begin{equation}\label{eq:Tucker}
\Theta=\tC\times_1\mM_1\times_1 \cdots \times_K \mM_K,
\end{equation}
where $\tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ is a core tensor, $\mM_k\in\mathbb{R}^{d_k\times r_k}$ are factor matrices with orthogonal columns, and $\times_k$ denotes the tensor-by-matrix multiplication~\cite{kolda2009tensor}. The Tucker low-rankness is popularly imposed in tensor analysis, and the rank determines the tradeoff between model complexity and model flexibility. Note that, unlike matrices, there are various notions of tensor low-rankness, such as CP rank~\cite{hitchcock1927expression} and train rank~\cite{oseledets2011tensor}. Some notions of low-rankness may lead to mathematically ill-posed optimization; for example, the best low CP-rank tensor approximation may not exist~\cite{de2008tensor}. We choose Tucker representation for well-posedness of optimization and easy interpretation.

The second constraint is that the entries of $\Theta$ are uniformly bounded in magnitude by a constant $\alpha \in \mathbb{R}_{+}$. In view of~\eqref{eq:latent}, we refer to $\alpha$ as the signal level. The boundedness assumption is a technical condition that avoids the degeneracy in probability estimation with ordinal observations.

@@

\section{Numerical Implementation}\label{sec:algorithm}
We describe the algorithm to seek the optimizer of~\eqref{eq:objective}. In practice, the cut-off points $\mb$ are often unknown, so we choose to maximize $\logl$ jointly over $(\Theta,\mb)\in\tP\times \tB$ (see the Supplement for details). The objective $\logl$ is concave in $(\Theta,\mb)$ whenever $f'$ is log-concave. However, the feasible set $\tP$ is non-convex, which makes the optimization~\eqref{eq:objective} a non-convex problem. We employ the alternating optimization approach by utilizing the Tucker representation of $\Theta$. Specifically, based on~\eqref{eq:Tucker} and~\eqref{eq:objective}, the objective function consists of $K+2$ blocks of variables, one for the cut-off points $\mb$, one for the core tensor $\tC$, and $K$ for the factor matrices $\mM_k$'s. The optimization is a simple convex problem if any $K+1$ out of the $K+2$ blocks are fixed. We update one block at a time while holding others fixed, and we alternate the optimization throughout the iteration. The convergence is guaranteed whenever $\logl$ is bounded from above, since the alternating procedure monotonically increases the objective. The Algorithm~\ref{alg} gives the full description.

@@

\begin{figure}[H]
\centering
\includegraphics[width=10cm]{algorithm.pdf}
\caption{Trajectory of objective function with various $d$ and $r$.}\label{fig:stability}
\end{figure}

Second, the algorithm takes the rank $\mr$ as an input. In practice, the rank $\mr$ is hardly known and needs to be estimated from the data. We use Bayesian information criterion (BIC) and choose the rank that minimizes BIC; i.e.,
\begin{align}\label{eq:BIC}
\hat \mr&=\argmin_{\mr\in\mathbb{N}^K_{+}} \text{BIC}(\mr)\\
&=\argmin_{\mr\in\mathbb{N}^K_{+}}\{-2\tL_{\tY}(\hat \Theta(\mr),\hat \mb(\mr))+p_e(\mr)\log (\prod_k d_k)\},
\end{align}
where $\hat \Theta(\mr), \hat\mb(\mr)$ are the estimates given the rank $\mr$, and $p_e(\mr)\stackrel{\text{def}}{=}\sum_k (d_k-r_k)r_k+\prod_k r_k$ is the effective number of parameters in the model. We select $\hat \mr$ that minimizes BIC through a grid search. The choice of BIC is intended to balance between the goodness-of-fit for the data and the degrees of freedom in the population model.

\section{Experiments}\label{sec:experiment}
In this section, we evaluate the empirical performance of our methods\footnote{Software package: \url{https://CRAN.R-project.org/package=tensorordinal}}. We investigate both the complete and the incomplete settings, and we compare the recovery accuracy with other tensor-based methods. Unless otherwise stated, the ordinal data tensors are generated from model~\eqref{eq:model} using standard probit link $f$. We consider the setting with $K=3$, $d_1=d_2=d_3=d$, and $r_1=r_2=r_3=r$. The parameter tensors are simulated based on~\eqref{eq:Tucker}, where the core tensor entries are i.i.d.\ drawn from $N(0,1)$, and the factors $\mM_k$ are uniformly sampled (with respect to Haar measure) from matrices with orthonormal columns. We set the cut-off points $b_\ell=f^{-1}(\ell/L)$ for $\ell\in[L]$, such that $f(b_\ell)$ are evenly spaced from 0 to 1. In each simulation study, we report the summary statistics across $n_{\text{sim}}$ = 30 replications.

@@

\begin{figure}[http]
\includegraphics[width=\textwidth]{panel.pdf}
\caption{Empirical relationship between (relative) MSE versus (a) dimension $d$, (b) signal level $\alpha$, (c) observation fraction $\rho$, and (d) number of ordinal levels $L$. In panels (b)-(d), we plot the relative MSE $=\FnormSize{}{\hat \Theta-\trueT}^2/\FnormSize{}{\trueT}^2$ for better visualization.}
\label{fig:finite}
\end{figure}

\subsection{Comparison with alternative methods}
Next, we compare our ordinal tensor method ({\bf Ordinal-T}) with three popular low-rank methods.
\begin{itemize}[itemsep=0pt,topsep=0pt,leftmargin=*,partopsep=0pt]
\item Continuous tensor decomposition ({\bf Continuous-T})~\cite{acar2010scalable} is a low-rank approximation method based on classical Tucker model.
\item One-bit tensor completion ({\bf 1bit-T})~\cite{ghadermarzy2018learning} is a max-norm penalized tensor learning method based on partial binary observations.
\item Ordinal matrix completion ({\bf Ordinal-M})~\cite{bhaskar2016probabilistic} is a rank-constrained matrix estimation method based on noisy, quantized observations.
\end{itemize}

We apply each of the above methods to $L$-level ordinal tensors $\tY$ generated from model~\eqref{eq:model}. The {\bf Continuous-T} is applied directly to $\tY$ by treating the $L$ levels as continuous observations. The {\bf Ordinal-M} is applied to the 1-mode unfolding matrix $\tY_{(1)}$. The {\bf 1bit-T} is applied to $\tY$ in two ways. The first approach, denoted {\bf 1bit-sign-T}, follows from~\citet{ghadermarzy2018learning} that transforms
$\tY$ to a binary tensor, by taking the entrywise sign of the mean-adjusted tensor, $\tY- |\Omega|^{-1}\sum_\omega y_\omega$. The second approach, denoted {\bf 1bit-category-T}, transforms the order-3 ordinal tensor $\tY$ to an order-4 binary tensor, $\tY^{\sharp}=\entry{y^{\sharp}_{ijkl}}$, via dummy variable encoding; i.e., $y^{\sharp}_{ijk\ell}=\mathds{1}\{y_{ijk}=\ell\}$ for $\ell\in[L-1]$. We evaluate the methods by their capabilities in predicting the most likely labels, $y_\omega^{\text{mode}}=\arg\max_\ell\mathbb{P}(y_\omega=\ell)$. Two performance metrics are considered: mean absolute deviation (MAD) $=d^{-K}\sum_\omega |y_\omega^{\text{mode}}-\hat y_\omega^{\text{mode}}|$, and misclassification rate (MCR) $=d^{-K}\sum_\omega\mathds{1}\{y_\omega^{\text{mode}}\neq\text{round}(\hat y_\omega^{\text{mode}})\}$,
where $\text{round}(\cdot)$ denotes the nearest integer. Note that MAD penalizes the large deviation more heavily than MCR.

@@

\begin{figure}[http]
\includegraphics[width=\textwidth]{panel2.pdf}
\caption{Performance comparison for predicting most likely labels. (a, c) Prediction errors versus sample complexity $\rho=|\Omega|/d^K$ when $L=5$. (b, d) Prediction errors versus the number of ordinal levels $L$ when $\rho=0.8$. }\label{fig:compare}
\end{figure}

\begin{figure}[http]
\includegraphics[width=\textwidth]{panel3.pdf}
\caption{Performance comparison for predicting median labels. (a, c) Prediction error versus sample complexity $\rho=|\Omega|/d^K$ when $L=5$. (b, d) Prediction error versus the number of ordinal levels $L$, when $\rho=0.8.$}\label{fig:compare2}
\end{figure}

We also compare the methods by their performance in predicting the median labels, $y_\omega^{\text{median}}=\min\{\ell\colon\mathbb{P}(y_\omega=\ell)\geq 0.5\}$. Under the latent variable model~\eqref{eq:latent} and Assumption~\ref{ass:link}, the median label is the quantized $\theta_\omega$ without noise; i.e.,\ $y_\omega^{\text{median}}=\sum_\ell \mathds{1}\{\theta_\omega\in(b_{\ell-1},b_\ell]\}$.
 We utilize the same simulation setting as in the earlier experiment. Figure~\ref{fig:compare2} shows that our method outperforms the others in both MCR and MAD. The improved accuracy comes from the incorporation of multilinear low-rank structure, multi-level observations, and the ordinal structure. Interestingly, the median estimator tends to yield smaller MAD than the mode estimator, $\text{MAD}(\tY^{\text{median}}, \hat \tY^{\text{median}})\leq \text{MAD}(\tY^{\text{mode}}, \hat \tY^{\text{mode}})$  (Figures~\ref{fig:compare}a-b vs.\ Figures~\ref{fig:compare2}a-b), for the three multilevel methods ({\bf 1bit-sign-T}, {\bf Ordinal-M}, and {\bf Ordinal-T}). The mode estimator, on the other hand, tends to yield smaller MCR than the median estimator, $\text{MCR}(\tY^{\text{mode}}, \hat \tY^{\text{mode}})\leq\text{MCR}(\tY^{\text{median}}, \hat \tY^{\text{median}})$ (Figures~\ref{fig:compare}c-d vs.\ Figures~\ref{fig:compare2}c-d). This tendency is from the property that the median estimator $\hat y^{(\text{median})}_\omega$ minimizes $R_1(z)=\mathbb{E}_{y_\omega}|y_\omega-z|$, whereas the mode estimator $\hat y^{(\text{mode})}_\omega $ minimizes $R_2(z)=\mathbb{E}_{y_\omega}\mathds{1}\{y_\omega=z\}$. Here the expectation is taken over the testing data $y_\omega$, conditional on the training data and thus parameters $\hat \Theta$ and~$\hat \mb$.

\section{Data Applications}\label{sec:dataapplication}
We apply our ordinal tensor method to two real-world datasets. In the first application, we use our model to analyze an ordinal tensor consisting of structural connectivities among 68 brain regions for 136 individuals from Human Connectome Project (HCP)~\citep{geddes2016human}. In the second application, we perform tensor completion to an ordinal dataset with missing values. The data tensor records the ratings of 139 songs on a scale of 1 to 5 from 42 users on 26 contexts~\citep{baltrunas2011incarmusic}.

\begin{table*}[http]
\includegraphics[width=1\textwidth]{HCPtable.pdf}
\caption{Node clusters in the HCP analysis. The first alphabet in the node name indicates the left (L) or right (R) hemisphere. The number in the parentheses indicates the node count in each cluster. }  \label{table:clustering}
\end{table*}

@@

\bibliography{tensor_wang}
\bibliographystyle{apalike}

\appendix
